---
title: "Problem Section 8"
subtitle:  "EmpiricalP-value"
graphics: yes
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(fastR2)
#Does this thing even work??
```

* * * 
### Instructions

Please answer the following questions in the order in which they are posed. Add a few empty lines below each and write your answers there. Focus on answering in complete sentences. You will also need scratch paper/pen to work out the answers before typing it.

For help with formatting documents in RMarkdown, please consult [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/). Another option is to search using Google. 

* * *

### Exercises 

1. Suppose the following sample - denoted by $X_1,X_2,\dots,X_n$ - are drawn from a $Norm(0, \sigma^2_0)$ distribution:
```{r}
x<- c(-0.58319935, -1.36090219,  0.38663763, -1.54365592,  0.87083945, -0.69187830, 0.45898841, -2.82556635,  0.01777137, -0.62753863 , 0.54611381, -1.39731591, -1.72584231,  0.91371529,  0.18096064, -0.53063107, -0.76604739, -1.97107704, 0.56394712,  1.13707563)
```

Suppose we want to test $H_0: \sigma^2_0 = 1$ versus $\sigma^2_0 > 1$ using the sample variance:
$$S^2 = \frac{1}{n-1} \sum\limits_{i=1}^{n} (X_i - \bar{X})^2.$$

a. Calculate the observed value for $S^2$. What direction is predicted for our statistic under $H_1$?

```{r 1a}
n = length(x)

(1/ (n - 1)) * sum( (x - mean(x))^2 )

var(x)
```

b. Simulate the null sampling distribution of $S^2$. Make a histogram, label it properly and mark the observed value with a vertical line. Calculate the empirical P-value.  (Hint: best to use a looping function such as `lapply`)

```{r label="gen-null-dist"}
set.seed(2626)
B <- 10000
sampsize <- length(x)
sigma2_null <- 1
```

c. Suppose we instead decide to the use the Range $R = X_{max} - X_{min}$ as our test statistic. Repeat parts a - c using $R$. (Write your code for c flexibly so you can calculate multiple statistics from the simulated data.)

d. Suppose we decide to use a level $\alpha=0.05$ test. What values of $S^2$ will you reject $H_0$ for? Same question for $R$. (Hint: find the 0.95 quantile of the null sampling distribution of each statistic. Think about why.) 

e. For the level $\alpha = 0.05$ test, compare the Type II error probabilities for $S^2$ and $R$ when $\sigma^2_0 = 3$. Remember, the Type II error rate is the probability that we fail to reject $H_0$ when $H_1$ is true. How do they do?   (You will need to simulate data under the alternative to calculate the Type II error rate)

```{r label="sim-alt-dist"}
set.seed(151)
sigma2_alt <- sqrt(3)

```
f. Think of a different statistic - call it $T$ - that will be sensitive to large values of $\sigma_0$. Repeat a - e using $T$. 
 



```{r eval = F}
#for Harshil: 
B <- 10000 
set.seed(1515)

null_sim_df <- lapply(X=1:B, FUN=function(X){
                           sampdata = rnorm(n=20,sd=1)
                           data.frame(s2 = var(sampdata), 
                                      r = diff(range(sampdata)))
})

null_sim_stats <- do.call(rbind, null_sim_df)

#find the 95th percentile of the null sampling distribution
critical_values <- null_sim_stats %>% 
                    summarise(s2crit = quantile(s2, 0.95),
                              rcrit = quantile(r,0.95))

sim_df_alt <- lapply(X=1:B, FUN=function(X){
                           sampdata = rnorm(n=20,sd=2)
                           data.frame(s2 = var(sampdata), 
                                      r = diff(range(sampdata)))
})

sim_stats_alt <- do.call(rbind, sim_df_alt)

sim_stats_alt %>% summarise( mean(s2 < critical_values$s2crit),
                             mean(r < critical_values$rcrit))


hist(sim_stats_alt$s2)

```
```{r}
set.seed(1414)
var(rexp(n=25, rate=2))
```

